Co-methylation

### 2015-04-17 Rmarkdown/comethylation.Rmd
Notes on estimating odds ratios and improved simulation code

>>>>

# Problem with LOR estimates

The simple estimator of within-fragment co-methylation is biased when the marginal probability of methylation is close to the boundary (0 or 1) unless the 2-tuple is sequenced at very high depth. I demonstrate this by simulating methylation patterns at 2-tuples as independent $binomail(1, p)$ random variables and showing that the simple LOR estimator is positively biased.

```{r}
# Function to simulate K 2-tuples with Poisson(mean_d) sequencing coverage and marginal
# probabilities p1, p2
simTables <- function(K, mean_d, p1, p2) {
  .simTables <- function(mean_d, p1, p2) {

    d <- rpois(1, mean_d)
    z1 <- factor(rbinom(d, 1, p1), levels = c(0, 1))
    z2 <- factor(rbinom(d, 1, p2), levels = c(0, 1))
    xtabs(~ z2 + z1, drop.unused.levels = FALSE)
  }
  replicate(K, .simTables(mean_d, p1, p2))
}

# Estimator of the (offset-adjusted) unconditional log odds ratios of K 2x2 tables
LOR <- function(x, offset = 0.5) {
  MM <- x[2, 2, ] + offset
  MU <- x[2, 1, ] + offset
  UM <- x[1, 2, ] + offset
  UU <- x[1, 1, ] + offset
  log(MM) + log(UU) - log(MU) - log(UM)
}

# Mantel-Haenszel estimator of a common log odds ratio of a 2x2xK table
MH <- function(x, offset = 0.5) {
  # TODO: return the standard error
  MM <- x[2, 2, ]
  MU <- x[2, 1, ]
  UM <- x[1, 2, ]
  UU <- x[1, 1, ]
  d <- apply(x, 3, sum)
  or_mh <- sum((UU * MM) / d) / sum((MU * UM) / d)
  log(or_mh)
}

# Estimator of the (offset-adjusted) unconditional log odds ratio of the 
# 2x2xK collapsed over K.
LORCollapsed <- function(x, offset) {
  z <- apply(x, c(1, 2), sum)
  # Artificially augment so it's a 2x2x1 table
  dim(z) <- c(dim(z), 1)
  LOR(z)
}

# Simulate and plot data
simAndPlot <- function(K, mean_d, p1, p2) {
  x <- mapply(function(p1, p2, K, mean_d) {
    simTables(K = K, mean_d = mean_d, p1 = p1, p2 = p2)
  }, p1 = p1, p2 = p2, MoreArgs = list(K = K, mean_d = mean_d), SIMPLIFY = FALSE)

  par(mfrow = c(3, 3))
  mapply(function(x, p1, p2) {
    hist(LOR(x), main = paste0("p1 = ", p1, ", p2 = ", p2), xlim = c(-8, 8), breaks = 16)
    abline(v = MH(x), col = "dodgerBlue", lwd = 3)
    abline(v = LORCollapsed(x), col = "orange", lwd = 3)
    abline(v = 0, col = "black", lty = 2)
  }, x = x, p1 = p1, p2 = p2)
  title(paste0("K = ", K, ", mean_d = ", mean_d), outer = TRUE, line = -1)
}

# Vary p1 with p2 = p1 at 30x coverage across 1000 CpG pairs
K <- 1000
mean_d <- 30
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
simAndPlot(K, mean_d, p1, p2)

# Vary p1 with p2 = 1 - p1 at 30x coverage across 1000 CpG pairs
K <- 1000
mean_d <- 30
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- 1 - p1
simAndPlot(K, mean_d, p1, p2)

# Vary p1 with p2 = p1 at 1000x coverage across 1000 CpG pairs
K <- 1000
mean_d <- 1000
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
simAndPlot(K, mean_d, p1, p2)

# TODO: p1 and p2 distributed U(0, 1) at 30x coverage across 1000 CpG pairs
# Will need to simulate p1 and p2 within .simTable().
```

## Following chat with Terry

These are my notes following a Skype conversion with Terry to discuss the issue.

- [ ] Try a higher order Taylor expansion of the log odds-ratio to get e.g., $n^{1}$ or $n^{3/2}$ convergence. The $+\frac{1}{2}$ fudge-factor is motivated by such an expansion and perhaps it is possible to do better by a higher order expansion.
  - [ ] First try showing the $c = \frac{1}{2}$ is optimal.
- [ ] Read texts about this issue
- [ ] Look for research papers on the general issue of dealing with bias in log odds-ratio estimates
- [ ] What if I just increase the fudge-factor to increase the shrinkage/smoothing of the estimates? It'll shrink to zero alright, but what is the theoretical justification?
- [ ] Check out the behaviour of the $\phi$ coefficient, i.e., Pearson's $r$ of the $2 \times 2$ table.
  - Landan et al.: "Note that correlation values are subject to considerable noise when coverage is low or when m1 or m2 are close to 0 or 1."
    - They don't define whether 'noise' means bias, variance, skewness, etc.

## Miscellaneous notes

Agresti in Categorical Data Analysis:

> In terms of bias and mean square error, Gart and Zweiful (1967) and Haldane (1955) showed the amended [i.e, add 0.5 to each count] estimators of the [odds-ratio log odds-ratio] behave well (see also Problem 12.4)

The Mantel-Haenszel estimator might be my best bet. Breslow (1981) says:

> The principal conclusion to be drawn from these results is that one should avoid using two of the standard techniques of odds ratio estimation with sparse data.

> The bias of the empirical logit estimator [i.e., + 0.5 to each cell], on the other hand, is sufficient erratic that the possibility of correction seems remote.

> Breslow (1981, Biometrika 68, 73-84) has shown that the Mantel-Haenszel odds ratio is a consistent estimator of a common odds ratio in sparse stratifications (Greenland, S. & Robins, J. M. Biometrics 41, 55–68 (1985).)

The Mantel-Haenszel estimator is an estimator of the common odds ratio for $2 \times 2 \times K$ tables. Therefore, using the MH estimator requires the assumption that all $2 \times 2$ tables have the same underlying odds ratio. The strata for co-methylation would be 'sample', 'IPD' and possibly 'region', where region is something like 'UMRs, LMRs, PMRs and other' or 'CGI and non-CGI'.

Some good notes on MH estimator [http://isites.harvard.edu/fs/docs/icb.topic79671.files/stratcch1.pdf](http://isites.harvard.edu/fs/docs/icb.topic79671.files/stratcch1.pdf).

An offset of 0.5 eliminates the first-order bias of the maximum likelihood estimator of the (log) odds ratio (Walter, S. D. Biostatistics (1987).)

Regarding the effect of the homogeneity on the Mantel-Haenszel estimator, Greenland, S. Stat Med 1, 217–227 (1982):

> Under the homogeneity assumption, OR^MH is an almost fully efficient estimator of the common value, and remains a consistent estimator of this value even with sparse data, i.e. when the fineness of stratification results in many small cell counts. __As shown later, its convergence properties under heterogeneity are not as desirable, although even in this case it remains insensitive to small cells.__ [emphasis added]

The _Miettenen_ estimator is like a generalised form of the Mantel-Haenszel estimator, but assumes each cell count is greater than approximately 4 (Greenland, S. Stat Med 1, 217–227 (1982)).
